\chapter{Background and Related Work}
\label{ch:background}
\raggedbottom


\section{WebAssembly and Stack-switching}
\label{sec:webassembly}

Stack-switching proposal introduces a type for continuation and instructions to work with continuations.
Continuation is an object that represents a function at some point in its execution. That point is either the beginning
of the function or a suspend instruction. Resuming continuation resumes the function from the recorded point.
Continuation can also be understood as \("\)the rest of a computation.\("\)

A good intuition for understanding resuming and suspending in the stack-switching proposal is that it is similar to
calling a function inside a try-catch block: the resume instruction (try block) specifies handlers (catch blocks) for
different suspension tags (exceptions). A function resumed via continuation can specify a suspension tag with the
suspend instruction. After that execution will jump to the corresponding handler. Here is where differences appear:
after jumping to the handler, there will be a new continuation object on the stack. Resuming this continuation will
resume the function that has just suspended.

In Figure~\ref{fig:stack-switching-example}
is an example of a code using stack-switching in Wasm.

To better understand what role each stack-switching primitive plays, let's look at TODO(stack-switching-primitives.tex).

TODO: make it into a list

\texttt{cont.new} creates a continuation from a function. Continuation has a type that reflects the signature of the
underlying function.

\texttt{resume} instruction resumes a continuation and
specifies what to do when the underlying function suspends. Depending on the type of the continuation, the
\texttt{resume} instruction also takes some values from the stack and passes them as arguments to the underlying
function.

\texttt{suspend} instruction suspends the execution. Suspension tags also have a type. The suspension tag type
determines what objects (other than the new continuation) will be on the stack at the handler site and the input type
of the new continuation(they don't have to be the same). This makes it possible to pass values in both directions:
from the main context into continuation and from continuation into the main context.

\begin{figure}[H]
    \centering
        \begin{lstlisting}[language=wasm, label={lst:wasm-example}, basicstyle=\tiny\ttfamily]
(func $generator
  ...
  (loop $loop
    ;; Suspend execution, pass current value of $i to consumer.
    (suspend $generate (local.get $i))
    ...
  )
)

(func $consumer
 ...
 ;; Create continuation executing function $generator.
 (local.set $generator_cont (cont.new $cont_type (ref.func $generator)))

 (loop $loop
   (block $on_generate (result i32 (ref $cont_type))
     ;; Resume continuation $generator_cont.
     ;; When $generate tag is thrown,
     ;; go to label $on_generate (points to right after this block)
     (resume $cont_type (on $generate $on_generate) (local.get $generator_cont))
     ;; $generator returned: no more data.
     (return)
   )
   ;; $generator suspended.
   ;; The stack now contains the value
   ;; produced by the generator and the new continuation.
   ;; Save continuation to resume it in the next iteration
   (local.set $generator_cont)
   ...
 )
)
        \end{lstlisting}
    \caption{Stack-switching example in WebAssembly}
    \label{fig:stack-switching-example}
\end{figure}




\section{Kotlin Coroutines}
\label{sec:kotlin-coroutines}

Kotlin provides several low-level intrinsics and primitives for implementing coroutines. These primitives form the
foundation for higher-level concurrency abstractions such as the kotlinx.coroutines library. The key primitives are:

\subsection*{Continuation}

\texttt{Continuation<T>}
is an interface that represents a continuation, similar to the continuation concept in WebAssembly described earlier. It
represents a coroutine at a suspension point - a place where the coroutine has paused its execution. The interface has
two main components:

\begin{itemize}
    \item \texttt{context: CoroutineContext} - the context of the coroutine
    \item \texttt{resumeWith(result: Result<T>)}
    - resumes the coroutine execution, passing a result value that becomes the return value of the suspension point
\end{itemize}

\subsection*{startCoroutineUninterceptedOrReturn}

This intrinsic function starts a coroutine. It is defined as an extension function on suspending function types:

\begin{lstlisting}[language=Kotlin, basicstyle=\small\ttfamily]
(suspend () -> T).startCoroutineUninterceptedOrReturn(
    completion: Continuation<T>
): Any?
\end{lstlisting}

The function immediately starts the suspending function with the given \texttt{completion} continuation. The
\texttt{completion}
parameter is the continuation that will be called when the coroutine finishes its execution (either with a result or an
exception). The function returns:
\begin{itemize}
    \item \texttt{COROUTINE\_SUSPENDED} if the coroutine was suspended
    \item The result value if the coroutine completed immediately without suspension
\end{itemize}

This primitive is used when you need direct control over coroutine execution.

\subsection*{createCoroutineUnintercepted}

This intrinsic creates a coroutine. It is defined as an extension function on suspending function types:

\begin{lstlisting}[language=Kotlin, basicstyle=\small\ttfamily]
(suspend () -> T).createCoroutineUnintercepted(
    completion: Continuation<T>
): Continuation<Unit>
\end{lstlisting}

Unlike \texttt{startCoroutineUninterceptedOrReturn}
, this function does not start the coroutine immediately. Instead, it returns a \texttt{Continuation<Unit>}
that can be resumed later to start the coroutine execution. This allows you to control when the coroutine begins.

\subsection*{suspendCoroutineUninterceptedOrReturn}

This intrinsic function gives you access to the current continuation inside a suspend function. It allows the function
to control whether it suspends or returns immediately:

\begin{lstlisting}[language=Kotlin, basicstyle=\small\ttfamily]
suspendCoroutineUninterceptedOrReturn<T>(
    block: (Continuation<T>) -> Any?
): T
\end{lstlisting}

The \texttt{block} receives the current continuation and can:
\begin{itemize}
    \item Return \texttt{COROUTINE\_SUSPENDED} to indicate that the coroutine has suspended and will be resumed later
    \item Return a value directly to complete immediately without suspension
\end{itemize}

This is the most basic building block for implementing custom suspending functions, as it gives you direct access to the
continuation and control over suspension behavior.
