\chapter{Conclusion}
\label{ch:conclusion}

This thesis presented an alternative implementation of Kotlin Coroutines for Kotlin/Wasm based on the
WebAssembly Stack-switching proposal. The implementation maps Kotlin coroutine primitives to
Wasm Stack-switching instructions, replacing the state-machine transformation approach used by the
existing Kotlin/JS-based implementation.

The evaluation showed measurable improvements in binary size: approximately 10.8\% reduction in production builds and
2.3\% in development builds for a small Compose Multiplatform project. On the other hand, performance benchmarks showed
a noticeable regression. The benchmark cases most representative of real-world code are around 2--2.5 times slower than the
baseline in V8. Cases with intensive suspension chaining regress by a factor of 4.5--5.7 without the
\btexttt{--experimental-wasm-growable-stacks} flag, and around 1.7 times with it.

Some of the observed performance regressions may decrease or even disappear as both the V8 Stack-switching
implementation and the Kotlin compiler support for it are developed further.

\section*{Current state}

The development of Kotlin Coroutines with Stack-switching is continued by JetBrains.
At the moment of writing this, the implementation is being prepared for
release as an experimental feature under a compiler flag. The production
implementation is based on the second implementation idea, which uses the immediate
resume optimization.

There are some benchmark results available for the production version of the implementation.
For example, KotlinConfApp~\cite{kotlinconf-app} was used for the size benchmark of production implementation.
Compiling it with the Stack-switching implementation gave around a 5\% improvement in binary size, which is consistent
with the analysis of the size benchmark presented in this thesis. Low-level performance benchmarks on the production
implementation also show a noticeable regression in performance, comparable to the previously presented performance
benchmark results. There are also higher-level Compose Multiplatform benchmarks available: those benchmarks show
performance results comparable to the baseline, being slower on some test cases and
faster on others. These results were presented at a meeting of the Wasm Community
Group~\cite{wasm-stack-meeting-2026}.

JetBrains is actively communicating with the V8 team regarding
the observed regressions from the Stack-switching implementation~\cite{wasm-stack-switching-repo}. The V8 team has
indicated that their Stack-switching implementation is still in active development, and further performance
improvements may follow.

\section*{Future work}

Several directions for future work follow from this thesis. First, the current implementation allocates a new
\btexttt{WasmContinuation} object for each suspension point. Reusing or pooling these objects could reduce garbage
collection pressure and improve performance. Second, a hybrid compilation strategy could be explored: using the
state-machine approach for some functions and stack-switching for others. Finally, once other production WebAssembly
engines add sufficient support for the Stack-switching proposal, benchmarks should be repeated on those engines to
determine whether the performance regression is specific to the current V8 implementation.
