\chapter{Evaluation}
\label{ch:evaluation}

\section{Correctness}
\label{sec:testing}

Kotlin codebase already contains extensive tests for coroutines. We can separate them into
two levels.

\subsection*{Intrinsic semantics tests}

These tests verify the behavior of the lowest-level coroutine
primitives like \startCoroutineUninterceptedOrReturn and \suspendCoroutineUninterceptedOrReturn.
They also cover mid-level coroutine APIs built directly on top of them like \texttt{startCoroutine}
and \texttt{suspendCoroutine}.

The tests are located in the Kotlin repository at
\btexttt{compiler/testData/codegen/box/coroutines/intrinsicSemantics}.

You can find an example of what kind of tests are in the \texttt{intrinsicSemantics} folder in Listing~\ref{lst:intrinsic-semantics-example} in the Appendix~\ref{ch:appendix}).

All discussed Stack-switching implementations pass all intrinsic semantics tests.

\subsection*{Box tests for coroutines}

This is an extensive collection of tests that verifies that coroutines work correctly when used
together with other Kotlin features.
Some of the things that are tested are:
\begin{itemize}
    \item \textbf{Control flow} (tests in \texttt{controlFlow/} directory): testing coroutines
    with \texttt{if}, \texttt{while}, \texttt{try-catch-finally}, and other control flow structures.
    \item \textbf{Default and variadic parameters} (e.g., \btexttt{defaultParametersInSuspend.kt},
    \btexttt{varargCallFromSuspend.kt}): testing suspend functions with default or variadic arguments.
    \item \textbf{Exception handling} (e.g., \texttt{handleException.kt}, \texttt{simpleException.kt}): testing
    that exceptions thrown inside coroutines are correctly passed to the completion continuation or
    handled by \texttt{try-catch} blocks.
\end{itemize}

As an example, below is a short test that verifies coroutine behavior with an \texttt{if} expression:

\begin{lstlisting}[language=Kotlin, caption={Example from controlFlow\_if1.kt}]
suspend fun s1(): Int = suspendCoroutineUninterceptedOrReturn { x ->
    x.resume(42)
    COROUTINE_SUSPENDED
}

fun f1(): Int = 117
fun f2(): Int = 1
fun f3(x: Int, y: Int): Int = x + y

fun box(): String {
    var result = 0
    suspend {
        result = f3(if (f1() > 100) s1() else f2(), 42)
    }.startCoroutine(object : Continuation<Unit> {
        override val context = EmptyCoroutineContext
        override fun resumeWith(result: Result<Unit>) {}
    })

    if (result != 84) return "fail: $result"
    return "OK"
}
\end{lstlisting}

The tests are located in the Kotlin repository at \btexttt{compiler/testData/codegen/box/coroutines}.

Both the initial and the optimized implementations pass 517 out of 523 coroutine tests. The 6 failing tests are
caused by a bug in the implementation: in certain cases, the generated Wasm code omits explicit casts to
\btexttt{Any?}. In Kotlin, all types are subtypes of \btexttt{Any?}, so no cast is needed; in Wasm, however,
such casts must be emitted explicitly. Adding the missing casts fixes the failures. These tests do not exercise
the coroutine suspension or resumption paths that are relevant to the size and performance analysis, so the
failures do not affect the conclusions of the evaluation.

\section{Size benchmark}
\label{sec:size-bench}

To evaluate the impact of the Stack-switching implementation on binary size, a small Kotlin project generated by
the Kotlin Multiplatform Wizard~\cite{kotlin-multiplatform-wizard}
was compiled using three different compiler versions:

\begin{itemize}
    \item \textbf{Baseline}: Kotlin 2.3.10 (original implementation without Stack-switching)
    \item \textbf{Stack-switching}: Base Stack-switching implementation rebased to Kotlin 2.3.10
    \item \textbf{Optimized SS}: Stack-switching with immediate resume optimization rebased to Kotlin 2.3.10
\end{itemize}

For each compiler version, two distributions were generated by running the following Gradle tasks:
\begin{itemize}
    \item \btexttt{wasmJsBrowserDevelopmentExecutableDistribution}: development build without optimizations
    \item \btexttt{wasmJsBrowserDistribution}: production build with optimizations
\end{itemize}

Table~\ref{tab:size-benchmark} shows the size of the compiled \btexttt{KotlinProject-composeApp.wasm} file in kilobytes
for each configuration.

\begin{table}[H]
\centering
\caption{Wasm binary size comparison (in KB)}
\label{tab:size-benchmark}
\begin{tabular}{|l|r|r|r|}
\hline
\textbf{Build Mode} & \textbf{Baseline} & \textbf{Stack-switching} & \textbf{Resume opt} \\
\hline
Dev & 20,460.67 & 19,987.97 & 19,992.46 \\
Prod & 2,082.82 & 1,857.17 & 1,858.34 \\
\hline
\end{tabular}
\end{table}

Table~\ref{tab:size-percentage} shows the percentage change in binary size compared to the baseline. Negative values indicate a size reduction.

\begin{table}[H]
\centering
\caption{Size change compared to baseline (negative values indicate reduction)}
\label{tab:size-percentage}
\begin{tabular}{|l|r|r|r|}
\hline
\textbf{Build Mode} & \textbf{Baseline} & \textbf{Stack-switching} & \textbf{Resume opt} \\
\hline
Dev & 20,460.67 KB & -2.3\% & -2.3\% \\
Prod & 2,082.82 KB & -10.8\% & -10.8\% \\
\hline
\end{tabular}
\end{table}

The results show that both Stack-switching implementations achieve a reduction in binary size compared to the
baseline. The size reduction is bigger in optimized builds (approximately 10.8\%) compared to development builds
(approximately 2.3\%). The difference between the base Stack-switching implementation and the optimized version is
negligible in terms of binary size, with both achieving similar reductions.

\section{Performance benchmarks}
\label{sec:performance-bench}

The performance of the implementation was measured by running benchmarks on low-level coroutine primitives.
The following cases are measured:

\begin{itemize}
    \item \btexttt{chainSuspensions.kt}: 20 coroutines are created and suspended. When the first coroutine is resumed, it
    suspends again, and in the suspend lambda the second coroutine is resumed. The second coroutine suspends and
    resumes the third coroutine in its suspend lambda. This continues on until the last coroutine is resumed, which
    records the fact that it was reached in a global variable. After that, all coroutines that were suspended so far
    are resumed so that they terminate properly.
    \item \btexttt{deepSuspensions.kt}: same setup as in \btexttt{chainSuspensions.kt}, but when the last coroutine
    in the chain is resumed, it also suspends and resumes 20 times.
    \item \btexttt{multipleSuspensions1.kt}: create one coroutine and suspend and resume it 50 times. A pattern
    with immediate resume inside suspend lambda is used.
    \item \btexttt{multipleSuspensions2.kt}: same as \btexttt{multipleSuspensions1.kt}, but the immediate resume pattern
    is not used. For each suspension the continuation is saved and resumed from the main context.
    \item \texttt{sequence.kt}: this benchmark is an exception as it's not a benchmark on low-level primitives. It's a
    benchmark for a Kotlin sequence builder, which is built on top of coroutines. For this test a sequence builder
    is used to compute the first 20 numbers from the Fibonacci sequence.
    \item \btexttt{startCoroutinePerformance.kt}: start 100 coroutines with \texttt{startCoroutine}.
\end{itemize}

Four code versions were used for comparison:

\begin{itemize}
    \item \textbf{Baseline} (commit \texttt{359b8303}): The original implementation without Stack-switching
    \item \textbf{Stack-switching} (commit \texttt{96b5d449}): Initial Stack-switching implementation
    \item \textbf{Optimized SS} (commit \texttt{4ffd6205}): Stack-switching with immediate resume optimization
    \item \textbf{Opt + param} (commit \texttt{2f6a6ebb}): Optimized SS with \btexttt{--experimental-wasm-growable-stacks} flag passed to V8
\end{itemize}

The benchmarks were executed in two compilation modes:

\begin{itemize}
    \item \textbf{dev mode}: Development build without optimizations
    \item \textbf{dce mode}: Production build with dead code elimination
\end{itemize}

All benchmarks were run on a MacBook Pro with an Apple M2 Max chip (12-core: 8 performance + 4 efficiency
cores, 64\,GB RAM) running macOS. The V8 version used was 14.7.89.

Each benchmark was run for 50 iterations after 50 warmup iterations. Some iterations were removed
as outliers before computing the average: any measurement exceeding $Q_3 + 1.5 \times IQR$ (where
$Q_3$ is the third quartile and $IQR$ is the interquartile range) was excluded, as such spikes are
typically caused by garbage collection pauses rather than the code under test. The bar charts in this
chapter display the average value across the remaining iterations, with error bars showing the 95\% confidence
interval computed from the filtered sample using the $t$-distribution.

\subsection{V8}

V8 benchmark results are the most relevant, since V8 is the engine used by Chromium, and consequently Google Chrome,
which is the most widely used browser~\cite{statcounter-browser-market-share}.

\subsubsection*{Raw results}

Tables~\ref{tab:benchmark-dev} and~\ref{tab:benchmark-dce} show the raw benchmark results for dev and dce modes respectively.

\begin{table}[H]
\centering
\caption{Benchmark results in dev mode (average time in nanoseconds, V8 runtime)}
\label{tab:benchmark-dev}
\begin{tabular}{|l|r|r|r|r|}
\hline
\textbf{Test} & \textbf{Baseline} & \textbf{Stack-switching} & \textbf{Optimized SS} & \textbf{Opt+param} \\
\hline
chainSusp & 18{,}644 & 93{,}227 & 96{,}872 & 30{,}711 \\
deepSusp & 23{,}119 & 118{,}298 & 102{,}826 & 42{,}617 \\
multiSusp1 & 7{,}196 & 14{,}114 & 7{,}500 & 8{,}114 \\
multiSusp2 & 6{,}930 & 14{,}395 & 16{,}000 & 16{,}439 \\
sequence & 2{,}250 & 9{,}214 & 11{,}935 & 9{,}809 \\
startCor & 25{,}020 & 37{,}625 & 42{,}562 & 32{,}842 \\
\hline
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Benchmark results in dce mode (average time in nanoseconds, V8 runtime)}
\label{tab:benchmark-dce}
\begin{tabular}{|l|r|r|r|r|}
\hline
\textbf{Test} & \textbf{Baseline} & \textbf{Stack-switching} & \textbf{Optimized SS} & \textbf{Opt+param} \\
\hline
chainSusp & 18{,}109 & 102{,}787 & 103{,}000 & 30{,}262 \\
deepSusp & 22{,}761 & 101{,}364 & 105{,}979 & 37{,}674 \\
multiSusp1 & 6{,}947 & 13{,}447 & 7{,}043 & 7{,}163 \\
multiSusp2 & 6{,}718 & 15{,}111 & 16{,}156 & 16{,}744 \\
sequence & 2{,}458 & 9{,}178 & 10{,}286 & 9{,}718 \\
startCor & 22{,}809 & 33{,}956 & 33{,}326 & 33{,}043 \\
\hline
\end{tabular}
\end{table}

\subsubsection*{Percentage Changes}

Table~\ref{tab:benchmark-percentage} presents the percentage changes relative to the baseline. Positive values indicate
performance regression (slower execution), while negative values indicate improvement (faster execution).

\begin{table}[H]
\centering
\caption{Percentage change compared to baseline (V8)}
\label{tab:benchmark-percentage}
\begin{tabular}{|l|c|r|r|r|r|r|}
\hline
\textbf{Test} & \textbf{Mode} & \textbf{Baseline} & \textbf{Stack-switching} & \textbf{Optimized SS} & \textbf{Opt+param} \\
\hline
chainSusp & dev & 18,644 & +400.0\% & +419.6\% & +64.7\% \\
deepSusp & dev & 23,119 & +411.7\% & +344.8\% & +84.3\% \\
multiSusp1 & dev & 7,196 & +96.1\% & +4.2\% & +12.8\% \\
multiSusp2 & dev & 6,930 & +107.7\% & +130.9\% & +137.2\% \\
sequence & dev & 2,250 & +309.5\% & +430.4\% & +335.9\% \\
startCor & dev & 25,020 & +50.4\% & +70.1\% & +49.5\% \\
\hline
chainSusp & dce & 18,109 & +467.6\% & +468.8\% & +68.8\% \\
deepSusp & dce & 22,761 & +345.3\% & +365.6\% & +65.5\% \\
multiSusp1 & dce & 6,947 & +93.6\% & +1.4\% & +3.1\% \\
multiSusp2 & dce & 6,718 & +124.9\% & +140.5\% & +149.2\% \\
sequence & dce & 2,458 & +273.3\% & +318.4\% & +295.3\% \\
startCor & dce & 22,809 & +48.9\% & +46.1\% & +44.9\% \\
\hline
\end{tabular}
\end{table}

\subsubsection*{Visual Comparison}

Figures~\ref{fig:benchmark-dev} and~\ref{fig:benchmark-dce} provide a visual comparison of the benchmark results across
all three implementations.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=7pt,
    width=\textwidth,
    height=8cm,
    ylabel={Average time (nanoseconds)},
    xlabel={Benchmark test},
    symbolic x coords={chainSusp,deepSusp,multiSusp1,multiSusp2,sequence,startCor},
    xtick=data,
    x tick label style={rotate=45, anchor=east},
    legend style={at={(0.5,-0.4)}, anchor=north, legend columns=4},
    ymin=0,
    enlarge x limits=0.1,
]
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,18644) +- (0,491)
    (deepSusp,23119) +- (0,276)
    (multiSusp1,7196) +- (0,213)
    (multiSusp2,6930) +- (0,216)
    (sequence,2250) +- (0,127)
    (startCor,25020) +- (0,1014)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,93227) +- (0,1141)
    (deepSusp,118298) +- (0,3182)
    (multiSusp1,14114) +- (0,388)
    (multiSusp2,14395) +- (0,252)
    (sequence,9214) +- (0,243)
    (startCor,37625) +- (0,2457)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,96872) +- (0,2576)
    (deepSusp,102826) +- (0,1577)
    (multiSusp1,7500) +- (0,219)
    (multiSusp2,16000) +- (0,241)
    (sequence,11935) +- (0,908)
    (startCor,42562) +- (0,4273)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,30711) +- (0,455)
    (deepSusp,42617) +- (0,1117)
    (multiSusp1,8114) +- (0,322)
    (multiSusp2,16439) +- (0,199)
    (sequence,9809) +- (0,284)
    (startCor,37400) +- (0,3253)
  };
\legend{Baseline, Stack-switching, Optimized SS, Opt + param}
\end{axis}
\end{tikzpicture}
\caption{Benchmark comparison in dev mode, V8 runtime (error bars show 95\% CI)}
\label{fig:benchmark-dev}
\end{figure}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=7pt,
    width=\textwidth,
    height=8cm,
    ylabel={Average time (nanoseconds)},
    xlabel={Benchmark test},
    symbolic x coords={chainSusp,deepSusp,multiSusp1,multiSusp2,sequence,startCor},
    xtick=data,
    x tick label style={rotate=45, anchor=east},
    legend style={at={(0.5,-0.4)}, anchor=north, legend columns=4},
    ymin=0,
    enlarge x limits=0.1,
]
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,18109) +- (0,406)
    (deepSusp,22761) +- (0,294)
    (multiSusp1,6947) +- (0,75)
    (multiSusp2,6718) +- (0,149)
    (sequence,2458) +- (0,158)
    (startCor,22809) +- (0,625)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,102787) +- (0,2925)
    (deepSusp,101364) +- (0,2961)
    (multiSusp1,13447) +- (0,453)
    (multiSusp2,15111) +- (0,487)
    (sequence,9178) +- (0,250)
    (startCor,33956) +- (0,703)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,103000) +- (0,1820)
    (deepSusp,105979) +- (0,1592)
    (multiSusp1,7043) +- (0,279)
    (multiSusp2,16156) +- (0,255)
    (sequence,10286) +- (0,556)
    (startCor,33326) +- (0,612)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,30568) +- (0,658)
    (deepSusp,37674) +- (0,442)
    (multiSusp1,7163) +- (0,211)
    (multiSusp2,16744) +- (0,190)
    (sequence,9718) +- (0,149)
    (startCor,33043) +- (0,428)
  };
\legend{Baseline, Stack-switching, Optimized SS, Opt + param}
\end{axis}
\end{tikzpicture}
\caption{Benchmark comparison in dce mode, V8 runtime (error bars show 95\% CI)}
\label{fig:benchmark-dce}
\end{figure}

\subsection{Reference Interpreter}

The Reference Interpreter~\cite{wasm-stack-switching-repo} is a reference interpreter with
a proof-of-concept implementation of Stack-switching. It is not designed to be fast, and performance benchmarks
on it are of limited interest as it is not used as a production VM. However, it is the VM that was initially
used for testing, and at that time no production VMs were available to run benchmarks.
The results are included here for completeness.

The following revisions were used:
\begin{itemize}
    \item \textbf{Baseline}: commit \texttt{0dbc8c49}
    \item \textbf{Stack-switching}: commit \texttt{01ae026e}
    \item \textbf{Optimized SS}: commit \texttt{c6c7bc2e}
\end{itemize}

\subsubsection*{Raw results}

Tables~\ref{tab:benchmark-ref-dev} and~\ref{tab:benchmark-ref-dce} show the raw benchmark results for dev and dce modes respectively.

\begin{table}[H]
\centering
\caption{Benchmark results in dev mode (average time in milliseconds, Reference Interpreter)}
\label{tab:benchmark-ref-dev}
\begin{tabular}{|l|r|r|r|}
\hline
\textbf{Test} & \textbf{Baseline} & \textbf{Stack-switching} & \textbf{Optimized SS} \\
\hline
chainSusp & 576 & 372 & 657 \\
deepSusp & 690 & 425 & 743 \\
multiSusp1 & 972 & 761 & 91 \\
multiSusp2 & 74 & 98 & 146 \\
sequence & 29 & 44 & 66 \\
startCor & 255 & 376 & 519 \\
\hline
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Benchmark results in dce mode (average time in milliseconds, Reference Interpreter)}
\label{tab:benchmark-ref-dce}
\begin{tabular}{|l|r|r|r|}
\hline
\textbf{Test} & \textbf{Baseline} & \textbf{Stack-switching} & \textbf{Optimized SS} \\
\hline
chainSusp & 519 & 297 & 554 \\
deepSusp & 601 & 330 & 629 \\
multiSusp1 & 926 & 709 & 61 \\
multiSusp2 & 48 & 60 & 93 \\
sequence & 18 & 26 & 40 \\
startCor & 152 & 218 & 310 \\
\hline
\end{tabular}
\end{table}

\subsubsection*{Percentage Changes}

Table~\ref{tab:benchmark-ref-percentage} presents the percentage changes relative to the baseline. Positive values indicate
performance regression (slower execution), while negative values indicate improvement (faster execution).

\begin{table}[H]
\centering
\caption{Percentage change compared to baseline (Reference Interpreter)}
\label{tab:benchmark-ref-percentage}
\begin{tabular}{|l|c|r|r|r|}
\hline
\textbf{Test} & \textbf{Mode} & \textbf{Baseline (ms)} & \textbf{Stack-switching} & \textbf{Optimized SS} \\
\hline
chainSusp & dev & 576 & $-35.5\%$ & $+14.1\%$ \\
deepSusp & dev & 690 & $-38.4\%$ & $+7.7\%$ \\
multiSusp1 & dev & 972 & $-21.6\%$ & $-90.6\%$ \\
multiSusp2 & dev & 74 & $+32.6\%$ & $+96.9\%$ \\
sequence & dev & 29 & $+51.0\%$ & $+124.4\%$ \\
startCor & dev & 255 & $+47.3\%$ & $+103.6\%$ \\
\hline
chainSusp & dce & 519 & $-42.9\%$ & $+6.7\%$ \\
deepSusp & dce & 601 & $-45.1\%$ & $+4.7\%$ \\
multiSusp1 & dce & 926 & $-23.4\%$ & $-93.4\%$ \\
multiSusp2 & dce & 48 & $+25.1\%$ & $+94.4\%$ \\
sequence & dce & 18 & $+44.7\%$ & $+119.5\%$ \\
startCor & dce & 152 & $+43.2\%$ & $+103.8\%$ \\
\hline
\end{tabular}
\end{table}

\subsubsection*{Visual Comparison}

Figures~\ref{fig:benchmark-ref-dev} and~\ref{fig:benchmark-ref-dce} provide a visual comparison of the benchmark results across
all three implementations.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=5pt,
    width=\textwidth,
    height=8cm,
    ylabel={Average time (milliseconds)},
    xlabel={Benchmark test},
    symbolic x coords={chainSusp,deepSusp,multiSusp1,multiSusp2,sequence,startCor},
    xtick=data,
    x tick label style={rotate=45, anchor=east},
    legend style={at={(0.5,-0.4)}, anchor=north, legend columns=3},
    ymin=0,
    enlarge x limits=0.1,
]
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,576) +- (0,2)
    (deepSusp,690) +- (0,3)
    (multiSusp1,972) +- (0,5)
    (multiSusp2,74) +- (0,0)
    (sequence,29) +- (0,0)
    (startCor,255) +- (0,7)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,372) +- (0,1)
    (deepSusp,425) +- (0,1)
    (multiSusp1,761) +- (0,4)
    (multiSusp2,98) +- (0,1)
    (sequence,44) +- (0,2)
    (startCor,376) +- (0,5)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,657) +- (0,2)
    (deepSusp,743) +- (0,2)
    (multiSusp1,91) +- (0,1)
    (multiSusp2,146) +- (0,1)
    (sequence,66) +- (0,0)
    (startCor,519) +- (0,1)
  };
\legend{Baseline, Stack-switching, Optimized SS}
\end{axis}
\end{tikzpicture}
\caption{Benchmark comparison in dev mode, Reference Interpreter (error bars show 95\% CI)}
\label{fig:benchmark-ref-dev}
\end{figure}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=5pt,
    width=\textwidth,
    height=8cm,
    ylabel={Average time (milliseconds)},
    xlabel={Benchmark test},
    symbolic x coords={chainSusp,deepSusp,multiSusp1,multiSusp2,sequence,startCor},
    xtick=data,
    x tick label style={rotate=45, anchor=east},
    legend style={at={(0.5,-0.4)}, anchor=north, legend columns=3},
    ymin=0,
    enlarge x limits=0.1,
]
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,519) +- (0,4)
    (deepSusp,601) +- (0,2)
    (multiSusp1,926) +- (0,4)
    (multiSusp2,48) +- (0,1)
    (sequence,18) +- (0,0)
    (startCor,152) +- (0,0)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,297) +- (0,1)
    (deepSusp,330) +- (0,2)
    (multiSusp1,709) +- (0,2)
    (multiSusp2,60) +- (0,0)
    (sequence,26) +- (0,0)
    (startCor,218) +- (0,3)
  };
\addplot+[error bars/.cd,y dir=both,y explicit]
  coordinates {
    (chainSusp,554) +- (0,3)
    (deepSusp,629) +- (0,3)
    (multiSusp1,61) +- (0,0)
    (multiSusp2,93) +- (0,0)
    (sequence,40) +- (0,0)
    (startCor,310) +- (0,2)
  };
\legend{Baseline, Stack-switching, Optimized SS}
\end{axis}
\end{tikzpicture}
\caption{Benchmark comparison in dce mode, Reference Interpreter (error bars show 95\% CI)}
\label{fig:benchmark-ref-dce}
\end{figure}

\subsection{Analysis}

The Stack-switching implementation is noticeably slower than the baseline on V8. The Reference Interpreter
results are less severe, but the Reference Interpreter is not a production VM and does not reflect the
experience of actual users.

\btexttt{chainSuspensions} and \btexttt{deepSuspensions} show the worst regression due to the nature of the V8
implementation, showing a 4.5--5.7 times slowdown. Passing the
\btexttt{--experimental-wasm-growable-stacks} flag reduces the regression to around 1.7 times
compared to the baseline, though other cases become slightly worse. Whether this flag is beneficial
depends on the particular application.

Among the presented benchmarks, \btexttt{multipleSuspensions2} is the closest to
what would appear in production code. This benchmark
is around 2--2.5 times slower than the baseline on V8, which is a noticeable regression.

The \btexttt{multipleSuspensions1} benchmark is comparable to the baseline, but the immediate resume pattern used in
this benchmark is very unlikely to appear in real code. Resuming a continuation immediately inside the
suspend lambda contradicts the purpose of suspending: if the resume value is available inside the
suspend lambda, the function does not need to suspend at all. The immediate-resume optimization therefore
has limited practical value.


\section{Summary}
\label{sec:evaluation-summary}

The Stack-switching implementation leads to a modest but noticeable decrease in binary size. The decrease is around 10\%
for a small application with little to no business logic written with Compose Multiplatform. The decrease would be
smaller for an application that has business logic not involving Kotlin Coroutines, since the parts of the
binary that do not use coroutines are not affected by the new implementation.

On the other hand,
performance benchmarks show a noticeable regression. The cases most similar to production code
are around 2--2.5 times slower. Cases with intensive suspension chaining are around 1.7 times slower if the
\btexttt{--experimental-wasm-growable-stacks} flag is passed to V8, and around 4.5--5.7 times slower without it. Cases with the
immediate resume pattern are comparable to the baseline, but they are unlikely to appear in production code.

\section{Threats to Validity}
\label{sec:threats}

Several factors limit the generalizability of the evaluation results.

\begin{itemize}
    \item \textbf{Limited number of applications.} The size benchmark uses a single small Compose Multiplatform
    project generated by the Kotlin Multiplatform Wizard. The result may not generalize to larger or more
    complex applications.
    \item \textbf{Immature V8 Stack-switching implementation.} The V8 Stack-switching support is still under
    active development. The observed performance regression may decrease as the implementation matures.
    \item \textbf{Low-level benchmarks.} The performance benchmarks focus on low-level coroutine primitives.
    They may not reflect the performance characteristics of higher-level application code.
    \item \textbf{Single VM.} Production benchmarks were only run on V8. Other production WebAssembly engines
    could not be tested due to incomplete or incompatible Stack-switching support.
    \item \textbf{Correctness test failures.} Six out of 523 coroutine tests fail due to an implementation bug
    (missing \btexttt{Any?} casts). While this bug does not affect the benchmarks, it indicates that the
    implementation is not fully complete.
\end{itemize}

\section{Note on other VMs}

No benchmarks for other production VMs are included because those VMs lack features necessary for the
Kotlin Coroutines implementation to function. The following is a list of VMs that have some progress in implementing
the Stack-switching proposal, but were not tested due to various issues.

\begin{itemize}
    \item \textbf{Wasmer}: according to the Wasm feature roadmap~\cite{wasm-features}, Wasmer fully supports
    Stack-switching. Unfortunately, it doesn't support another Wasm feature that is necessary for Kotlin/Wasm as a
    whole: Garbage collection.
    \item \textbf{SpiderMonkey (Mozilla Firefox)}: Stack-switching is only partially implemented. The implemented
    feature subset is not enough to run the Kotlin Coroutines implementation. While adapting the implementation to run
    only on the supported primitives is technically possible, it is not straightforward and the result would be substantially
    different from the initial implementation~\cite{spidermonkey-stack-switching}.
    \item \textbf{Wasmtime}: Stack-switching implementation in Wasmtime has a bug which prevents Kotlin Coroutines from running properly~\cite{wasmtime-stack-switching-bug}.
\end{itemize}

One more VM that could technically be tested is Wizard~\cite{wizard-engine}. It has support for Stack-switching, but like
Reference Interpreter, it is not a production VM, so the benchmark results would be of limited interest.