\chapter{Implementation}
\label{ch:implementation}

To implement Kotlin coroutines using Wasm stack-switching proposal we have to implement the intrinsics and primitives
discussed in the previous section. Kotlin coroutine primitives can be mapped to the Wasm stack-switching
primitives pretty closely:
\begin{itemize}
    \item \textbf{Continuation} maps to Wasm continuation objects.

    \item \textbf{createCoroutineUnintercepted} maps to \texttt{cont.new}.

    \item \textbf{startCoroutineUninterceptedOrReturn} maps to \texttt{cont.new} followed by \texttt{resume}.

    \item \textbf{suspendCoroutineUninterceptedOrReturn} maps to the \texttt{suspend} instruction with some additional
    logic to execute suspend lambda (the \texttt{block} argument that is passed to \texttt{suspendCoroutineUninterceptedOrReturn}).
    \item \textbf{Continuation.resumeWith} maps to the \texttt{resume} instruction.
\end{itemize}

This correspondence can be observed more closely on figure TODO(figure that puts coroutine-primitives.tex and stack-switching-primitives.tex side by side)

There is one unfortunate difference in how suspension is handled in Kotlin, which makes it not so obvious how to
implement it in Wasm. In Kotlin the function \texttt{suspendCoroutineUninterceptedOrReturn}
accepts a lambda which either returns some result or suspends the execution. This lambda accepts an argument: current
continuation. Resuming this continuation will resume the function. The intention of providing this argument is
for the suspension lambda to contain code that saves the continuation in some variable. Then, in case
suspension really happens, the function will be resumed by resuming this saved continuation.

For example, consider this simple suspend function:

\begin{lstlisting}[language=Kotlin, basicstyle=\footnotesize\ttfamily]
var savedContinuation: Continuation<String>? = null

suspend fun suspendHere(): String = 
    suspendCoroutineUninterceptedOrReturn { cont ->
        // Save continuation before suspending
        savedContinuation = cont
        // Indicate suspension occurred
        COROUTINE_SUSPENDED
    }
\end{lstlisting}

In this code, the continuation \texttt{cont} is available inside the lambda and can be saved to
\texttt{savedContinuation} before the function actually suspends. Later, calling
\texttt{savedContinuation?.resume("result")} will resume the suspended function.

However, there is a more subtle case that must also be supported: the continuation can be resumed immediately inside
the suspension lambda itself. Consider this example:

\begin{lstlisting}[language=Kotlin, basicstyle=\footnotesize\ttfamily]
suspend fun suspendFun(): String =
    suspendCoroutineUninterceptedOrReturn { cont ->
        cont.resume("OK")      // Resume immediately
        COROUTINE_SUSPENDED    // Still indicate suspension
    }

fun test() {
    var completionResult: String? = null
    val result = suspendFun.startCoroutineUninterceptedOrReturn(
        object : Continuation<String> {
            override val context = EmptyCoroutineContext
            override fun resumeWith(value: Result<String>) {
                completionResult = value.getOrThrow()
            }
        }
    )
    // result == COROUTINE_SUSPENDED
    // completionResult == "OK"
}
\end{lstlisting}

In this case, even though the continuation is resumed immediately inside the lambda, the function still returns
\texttt{COROUTINE\_SUSPENDED} to the caller. The actual result \texttt{"OK"} is delivered to the completion
continuation via its \texttt{resumeWith} method. This behavior is important because it maintains the invariant that
when \texttt{suspendCoroutineUninterceptedOrReturn} returns \texttt{COROUTINE\_SUSPENDED}, the caller should not
use the return value but instead wait for the completion continuation to be invoked.

Essentially, the problem is that in Kotlin the continuation object has to be available before we even know whether
a function suspends or not. On the contrary, in Wasm we only get the continuation object for the suspended function
after the function actually suspends, and we return to the call site. The following sections describe how to bridge
this gap.

\section{Base Implementation}
\label{sec:base-implementation}

The base implementation of Kotlin coroutines using Wasm stack-switching consists of several key components that work
together to bridge the gap between Kotlin's coroutine model and Wasm's continuation primitives.

\subsection{Core Implementation}
\label{subsec:core-intrinsics}

The implementation provides several compiler intrinsics that map directly or very closely to Wasm stack-switching
instructions:
\texttt{suspendIntrinsic} (maps to \texttt{suspend}), \texttt{suspendFunctionToContref} (uses \texttt{cont.new} and
\texttt{cont.bind}), and \texttt{resumeWithIntrinsic}/\texttt{resumeThrowIntrinsic} (use \texttt{resume} and
\texttt{resume\_throw} within handler blocks).

The core of the implementation is the \texttt{WasmContinuation} class, which extends \texttt{CoroutineImpl} and wraps
a Wasm continuation reference (\texttt{contref}). The \texttt{CoroutineImpl} class was originally implemented for the
state machine-based coroutine implementation and provides the foundation for managing coroutine state:

\begin{itemize}
    \item \textbf{State management}: \texttt{state} and \texttt{exceptionState} track the current execution state and
    exception handling state of the coroutine
    \item \textbf{Result storage}: \texttt{result} and \texttt{exception} store the current result or exception to
    pass when resuming
    \item \textbf{Suspension tracking}: \texttt{wasSuspended} flag indicates whether the coroutine has suspended at
    least once
    \item \textbf{Completion handling}: \texttt{resultContinuation} is the continuation to call when the coroutine
    completes
\end{itemize}

The \texttt{WasmContinuation} class adds Wasm-specific functionality:

\begin{itemize}
    \item \texttt{wasmContBox} -- a box containing the Wasm continuation reference, which can be updated as the
    coroutine suspends and resumes
    \item \texttt{isResumed} -- a flag to ensure each continuation instance is resumed only once
\end{itemize}

The key method is \texttt{doResume()}, which handles the mismatch between Kotlin's and Wasm's suspension behavior.
The \texttt{CoroutineImpl.resumeWith()} method calls \texttt{doResume()} and handles the result: if \texttt{doResume()}
returns \texttt{COROUTINE\_SUSPENDED}, the coroutine remains suspended; otherwise, it completes and invokes the
completion continuation with the result or exception.

The key challenge is that in Kotlin, \texttt{suspendCoroutineUninterceptedOrReturn} provides the continuation object to
the suspend lambda \emph{before} suspension occurs. However, in Wasm, the continuation object only becomes available
\emph{after} the \texttt{suspend} instruction executes and control returns to the caller. The implementation solves this
problem using a two-phase approach in \texttt{doResume()}:

\begin{enumerate}
    \item \textbf{Resume Phase}: When resuming a Wasm continuation, the resume intrinsic is called within a handler
    block. If the resumed function suspends again, the handler catches the suspension and returns a
    \texttt{ResumeIntrinsicResult} containing:
    \begin{itemize}
        \item \texttt{suspendBody} -- the lambda that was passed to \texttt{suspendCoroutineUninterceptedOrReturn}
        \item \texttt{remainingFunction} -- the new Wasm continuation created by the suspension
        \item \texttt{result} -- the value if the function completed without suspending
    \end{itemize}

    \item \textbf{Suspend Body Execution}: If a \texttt{suspendBody} is returned, it means the function suspended. The
    \texttt{doResume()} method then executes this suspend body, passing the \texttt{WasmContinuation} itself as the
    continuation argument. This gives the suspend lambda access to a continuation object that can be saved and resumed
    later.
\end{enumerate}

The \texttt{doResume()} method runs in a loop, repeatedly resuming the Wasm continuation and executing suspend bodies
until either:
\begin{itemize}
    \item The coroutine completes with a final result
    \item The coroutine suspends and the suspend body returns \texttt{COROUTINE\_SUSPENDED}, indicating that the
    continuation has been saved and will be resumed asynchronously
\end{itemize}

\section{Suspension count optimization}
\label{sec:optimization}

The base implementation described in the previous section handles all cases correctly, including the case where a
continuation is resumed immediately inside the suspend lambda. However, this case presents an optimization opportunity.
When a continuation is resumed immediately inside the suspend body, what does this actually represent? It represents
that we are resuming the current function right away. This is semantically equivalent to not suspending at all -- the
function can simply return the result directly to its caller without performing any Wasm stack-switching operations.

Consider again the example from earlier where the continuation is resumed immediately:

\begin{lstlisting}[language=Kotlin, basicstyle=\footnotesize\ttfamily]
suspend fun suspendFun(): String =
    suspendCoroutineUninterceptedOrReturn { cont ->
        cont.resume("OK")      // Resume immediately
        COROUTINE_SUSPENDED    // Still indicate suspension
    }
\end{lstlisting}

In the base implementation, this code would:
\begin{enumerate}
    \item Execute the Wasm \texttt{suspend} instruction to suspend the function
    \item Return control to the caller with the Wasm continuation
    \item The caller would then resume this continuation with the result
    \item The Wasm \texttt{resume} instruction would restart the suspended function
\end{enumerate}

However, since the continuation is resumed immediately, we can optimize this to simply return the value directly,
avoiding the expensive suspend and resume operations entirely.

\subsection{Implementation Approach}

The optimization is implemented using two key mechanisms: synthetic continuations and suspension tracking.

\subsubsection{Synthetic Continuations}

When \texttt{suspendCoroutineUninterceptedOrReturn} is called, instead of immediately executing the Wasm
\texttt{suspend} instruction, the implementation creates a \emph{synthetic} \texttt{WasmContinuation} object. This
synthetic continuation is not backed by an actual Wasm continuation reference initially (its \texttt{contref} is null).
The synthetic continuation is passed to the suspend lambda.

If the suspend lambda resumes this synthetic continuation immediately by calling \texttt{resumeWith}, the continuation
stores the result in a special field (\texttt{syntheticResumeValue}) instead of performing an actual Wasm resume
operation. After the suspend lambda returns, \texttt{suspendCoroutineUninterceptedOrReturn} checks whether the
synthetic continuation was resumed. If it was, the function returns the stored value directly without calling the
\texttt{suspendIntrinsic} (which would execute the Wasm \texttt{suspend} instruction).

\subsubsection{Suspension Tracking}

To maintain correct semantics, the implementation must track whether a real suspension occurred. This is necessary
because the completion continuation should only be invoked once, and only when a real suspension happens. The
implementation uses a \texttt{SuspensionMarker} context element that is added to the coroutine context. This marker
contains a \texttt{wasSuspended} flag that is set to true when \texttt{suspendCoroutineUninterceptedOrReturn} returns
\texttt{COROUTINE\_SUSPENDED} without an immediate resume.

When the coroutine completes, the \texttt{CoroutineImpl.resumeWith} method checks both the coroutine's internal
\texttt{wasSuspended} flag and the \texttt{SuspensionMarker} to determine whether a real suspension occurred. If no
real suspension occurred (immediate resume case), the method prevents double-completion by returning early instead of
invoking the completion continuation again.

\subsection{Benefits}

This optimization eliminates unnecessary Wasm stack-switching operations in cases where the continuation is resumed
immediately. Since stack-switching operations involve saving and restoring execution state, avoiding them when possible
improves performance. The optimization is beneficial for coroutines that frequently use immediate
resumption patterns, such as those that check cached values or perform fast-path optimizations before potentially
suspending.

\section{Testing and Validation}
\label{sec:testing}

\section{Summary}
\label{sec:implementation-summary}
