\chapter{Introduction}
\label{ch:introduction}

\section{Context}

WebAssembly (Wasm) is a binary instruction format for a stack-based virtual machine~\cite{wasm-spec}. Its most mainstream use case is in
browsers. Many languages can be compiled to it, including Kotlin. Kotlin/Wasm is used, for example, by the Compose
Multiplatform library~\cite{compose-multiplatform} for its Web target. Compose Multiplatform also uses Kotlin Coroutines extensively in its
implementation.

The current implementation of Kotlin Coroutines for Kotlin/Wasm is based on the implementation for Kotlin/JS, which,
similar to Kotlin/JVM, uses state-machine transformation on suspend functions. Each suspend function is compiled into
a state machine. This transformation adds generated code to the binary and introduces overhead at runtime.

Wasm has a proposal called Stack-switching. It introduces new primitives for suspending and resuming execution
contexts at the Wasm level, which can be used to write an alternative implementation of Kotlin Coroutines for
Kotlin/Wasm without state-machine transformation.

\section{Goal}
\label{sec:motivation}

The goal of this thesis is to implement Kotlin Coroutines for Kotlin/Wasm using WebAssembly Stack-switching
primitives, and to evaluate whether this implementation improves binary size and performance compared to the
existing state-machine-based approach.

A central technical challenge is the mismatch between Kotlin's coroutine model, where a continuation object must be
available before a suspend function actually suspends, and the Wasm Stack-switching model, where the continuation
only becomes available after suspension. This thesis describes two implementations that address this mismatch.

The main findings are: the Stack-switching implementation reduces binary size (by approximately 10.8\% in production
builds for a small Compose Multiplatform project), but currently shows a performance regression compared to the
baseline on V8, the JavaScript and WebAssembly engine used by Chromium-based browsers.

\section{Thesis Structure}
\label{sec:structure}

The thesis is organized as follows: Chapter~\ref{ch:background} provides background information on WebAssembly
Stack-switching and Kotlin coroutines primitives. Chapter~\ref{ch:implementation} describes the implementation of
Kotlin coroutines using the WebAssembly Stack-switching proposal. Chapter~\ref{ch:evaluation} evaluates the
implementation in terms of correctness, binary size, and performance. Finally, Chapter~\ref{ch:conclusion} concludes the thesis and
discusses the current state of the coroutines with Stack-switching implementation.
