\chapter{Related Work}
\label{ch:related-work}
\raggedbottom

This chapter discusses approaches related to compiling coroutines and asynchronous code to WebAssembly,
as well as alternative designs for Stack-switching in Wasm.

\section{Approaches to Asynchronous Code in WebAssembly}

\subsection{Asyncify}

Before native stack-switching support, one of the most widely used techniques for supporting asynchronous-style code in
WebAssembly was Asyncify~\cite{asyncify}, a binary transformation pass implemented in Binaryen.
Asyncify works by rewriting a compiled Wasm binary to make it possible to pause and resume execution at
arbitrary points. At suspension points, the transformation inserts code to unwind the current call stack
into a pre-allocated buffer, and at resumption points it inserts code to rewind the stack from that buffer.

The approach is analogous in motivation to the state-machine transformation used by the Kotlin compiler for the JVM and JS
targets: both enable suspension and resumption without native stack-switching support. However, Kotlin performs a
source/IR-level stackless coroutine lowering, whereas Asyncify instruments the generated Wasm binary to save and restore
stack state. The key difference is that Asyncify operates at the Wasm binary level rather than at the source or IR level,
making it language-agnostic.

Asyncify has been used by several language runtimes and ports, including Python (Pyodide) and some Ruby-on-Wasm efforts.
Its main drawbacks are increased binary size and runtime overhead from the stack save and restore
operations~\cite{asyncify}.

\subsection{JS Promise Integration (JSPI)}

An alternative approach for asynchronous WebAssembly is the JS Promise Integration (JSPI) proposal~\cite{jspi-proposal}.
JSPI provides a JavaScript API that allows Wasm imports and exports to be wrapped so that a
synchronous Wasm call can suspend waiting for a JavaScript \texttt{Promise} to settle, and then be resumed
when the promise resolves. From the Wasm binary's perspective, JSPI can be transparent: unlike Asyncify, it does not
require rewriting the module, although the relevant imports and exports must be wrapped by the JavaScript host.
The suspension and resumption are managed by the JavaScript engine.

JSPI addresses the specific use case of interoperating with asynchronous JavaScript APIs from synchronous
Wasm code. JSPI does not expose general-purpose stack-switching primitives within the Wasm module itself.
Therefore, by itself, it is not a suitable mechanism for implementing arbitrary language-level coroutines
that suspend and resume entirely within Wasm. The Stack-switching proposal is a more general solution that covers this use case as well.

JSPI has been presented and discussed in the WebAssembly Stack Subgroup~\cite{jspi-wasm-meeting}. It has also
been partially implemented in SpiderMonkey using Stack-switching primitives~\cite{spidermonkey-stack-switching-diff}.

\subsection{WasmFX and Typed Continuations}

WasmFX~\cite{wasmfx-proposal} is an alternative proposal for adding stack-switching capabilities to
WebAssembly, based on the theory of algebraic effects and typed continuations.
WasmFX also uses typed continuations, with core operations such as \texttt{cont.new}, \texttt{resume}, and \texttt{suspend},
and additional constructs such as \texttt{cont.bind}, \texttt{resume\_throw}, and \texttt{barrier}.
The core difference from the Stack-switching proposal is that WasmFX continuations are
designed to compose with effect handlers, enabling a richer set of control-flow abstractions.

WasmFX was presented and developed in the WebAssembly Stack Subgroup over several meetings~\cite{wasmfx-meeting-2021,wasmfx-meeting-2023}.
A prototype implementation in the Wasmtime engine was discussed in the subgroup in 2023~\cite{wasmfx-wasmtime-2023}.
The current Stack-switching proposal has evolved alongside WasmFX/typed-continuation work, and the two lines of work have influenced discussions in the Stack Subgroup.

\section{Other Language Runtimes on WebAssembly}

Multiple language communities have explored how to compile their runtimes to WebAssembly, which requires
solving problems similar to the ones addressed in this thesis.

\textbf{Go and TinyGo.} The Go runtime uses goroutines, lightweight runtime-managed threads with dynamically managed stacks. Two separate presentations
in the WebAssembly Stack Subgroup discussed the challenges of compiling Go to Wasm: one covering the full
Go runtime~\cite{go-wasm-meeting}, and one covering TinyGo~\cite{tinygo-wasm-meeting}, a Go compiler for
embedded and Wasm targets. Both noted the difficulty of mapping Go's concurrency model to WebAssembly's
linear stack model.

\textbf{OCaml.} Multicore OCaml adds support for algebraic effects, which generalize coroutines and other
control-flow abstractions. The OCaml community presented its experience compiling to Wasm in the
Stack Subgroup~\cite{ocaml-wasm-meeting}, discussing how native OCaml effects could be compiled to
WebAssembly using stack-switching primitives.

\textbf{Erlang.} Erlang's language model is based on lightweight isolated processes with their own execution state and message passing.
Challenges in implementing the Erlang runtime in WebAssembly were presented in the
Stack Subgroup~\cite{erlang-wasm-meeting}, noting that native Wasm stack switching could simplify such ports
compared with approaches based on binary instrumentation such as Asyncify.

These cases, together with Kotlin, illustrate that language-level concurrency abstractions and
stack-switching support in WebAssembly are broadly needed across many language communities.

